---
title: "Basic HSItools workflow"
vignette: >
  %\VignetteIndexEntry{basic_workflow}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Setup

```{r}
#| label: setup
library(HSItools)
```

# Reflectance

Typically, you need to calculate reflectance from raw data captured during the scan. HSItools makes this straightforward using a combination of {[terra](https://CRAN.R-project.org/package=terra)} and {HSItools} functions.

First, locate your data. Because no assumptions about directory structure are made, you can use external white and dark reference files in this step. This is useful when, for example, high exposure time during the scan led to overexposure of your white reference and you need to substitute an external one.

If you're reading data from ENVI .raw files, it may be important to use `noflip = TRUE` in your `terra::rast()` call, otherwise files will be flipped vertically.

Our test dataset is a tiny 9 by 9 pixel subset of HSI data from the Hyperspectral Imaging Laboratory in Gdańsk.

```{r}
# Read in captured data and references
data <- list(
  x = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/testdata.tif"
    )
  ),
  whiteref = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/WHITEREF_testdata.tif"
    )
  ),
  darkref = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/DARKREF_testdata.tif"
    )
  )
)
```

With this data you are ready to calculate reflectance, a fraction between 0 and 1 representing "complete" darkness and whiteness. You can control integration times for your specimen and white reference with the `tint` argument, which defaults to `c(1, 1)`, assuming both use the same integration time. If you are using external references with different integration times, adjust accordingly; the white reference integration time always comes first. Importantly, we're also scaling the dark reference values, which could be really important for lower singal to noise ratio (SNR) systems, such as SWIR cameras.

You can also control whether processing should be done in memory or not. The default produces as many temporary files as there are layers in the SpatRaster. If your system has sufficient RAM, you can speed up computation by processing in memory. Finally, you can write your result to file or keep it in memory. If you process on disk it makes sense to also write your result to disk. In this instance {\[withr\]https://cran.r-project.org/package=withr} takes care of temporary files, and removes them once function finishes. If you decide to calculate using temporary files and store result in memory, ordinary temp file is used and might not be cleaned automatically.

```{r}
# Calculate with defaults: temporary files, result in memory
r1 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref
)

# Calculate entirely in memory
r2 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
)

# Calculate in memory and write to file
r3 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE,
  filename = tempfile(fileext = ".tif")
)

# Results should be nearly identical, differing only at distant decimal places
terra::all.equal(r1, r2)
```

# Quantizing

High resolution (spatial and spectral) camera, especially VNIR, could produce large reflectance files, once we step out of DN (digital number) integer space into floating point numbers. This kind of data can be stored as a "scalled" or "quantized" version, where data is multiplied by a known factor allowing for storage of floats as integers. This reduces the filesize often by more than 50%. A tradeoff is lower precission of restored data. {terra} reads quantized data easily. This approach is suitable for cameras with high SNR. Do not attempt to quantize low SNR data like current generation SWIR sensors.

```{r}
r4 <- HSItools::hsi_write_scaled(
  x = r3,
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)
```

# Smoothing: spatial

Next, apply a spatial filter such as a simple median to remove outliers. The default window size is 3 by 3. As always, you can write to file.

```{r}
r5 <- HSItools::hsi_smooth_median(x = r4)
```

# Smoothing: spectral

With data smoothed across the spatial dimension, there should be no NAs in the spectral series. Now you can apply a Savitzky-Golay filter to smooth the reflectance spectrum in each pixel. Several arguments are available to control the smoothing parameters. The function uses the {[gsignal](https://CRAN.R-project.org/package=gsignal)} implementation.

```{r}
r6 <- HSItools::hsi_smooth_savgol(x = r5)
```

You might be interested also in calculating a derivatives of reflectance. It is easily done by changing the derivative order in the same function. For a 1st derivative, set m = 1.

```{r}
r_derivative <- HSItools::hsi_smooth_savgol(
  x = r5,
  m = 1
)
```

# Continuum removal

Smoothed data can be enhanced by removing the continuum. This corrects for the inherent tendency in spectral data to slope toward one edge of the sensing range. This is a computationally intensive function, so consider using it on subsets or regions of interest (ROI) rather than full-resolution data. The function uses the {[prospectr](https://CRAN.R-project.org/package=prospectr)} implementation.

```{r}
r7 <- r6 |>
  HSItools::hsi_remove_continuum()
```

# Workflow

With {HSItools} and {terra} it is easy to pipe operations into a single workflow. You can write to file at any point.

```{r}
# Calculate reflectance
r <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
) |>
  # Smooth spatially
  HSItools::hsi_smooth_median() |>
  # Smooth spectrally and write intermediate result
  HSItools::hsi_smooth_savgol(filename = tempfile(fileext = ".tif")) |>
  # Optionally remove continuum
  HSItools::hsi_remove_continuum()
```

# Tilled processing

If your computer offers multiple cores you can use this to apply pixel-wise operations in parallel. We're using lambda notation to pass the functions. You also need to specify the arguments.

::: {.callout-caution appearance="minimal"}
Tilled processing produces reliable results only for operations that work on a single pixel basis. Spatial operation, like median, should never be run this way.
:::

```{r}
library(mirai)
# Set number of parallel workers, never set it equal to max number of available cores. Use max n-1.
cores <- 2

mirai::daemons(cores)
```


```{r}
r <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
) |>
  # Smooth spatially - no tiling
  HSItools::hsi_smooth_median() |>
  # Smooth spectrally and write intermediate result - you can use tiling
  HSItools::hsi_tiled(
    fun = \(tile) HSItools::hsi_smooth_savgol(tile),
    n_tiles = cores,
    filename = tempfile(fileext = ".tif"),
    overwrite = TRUE
  ) |>
  HSItools::hsi_tiled(
    fun = \(tile) HSItools::hsi_remove_continuum(tile),
    n_tiles = cores,
    filename = tempfile(fileext = ".tif"),
    overwrite = TRUE
  )
```