---
title: "Basic HSItools workflow"
vignette: >
  %\VignetteIndexEntry{basic_workflow}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Setup

```{r}
#| label: setup
library(HSItools)
```

# Reflectance

Typically, you need to calculate reflectance from raw data captured during the scan. HSItools makes this straightforward using a combination of {[terra](https://CRAN.R-project.org/package=terra)} and {HSItools} functions.

First, locate your data. Because no assumptions about directory structure are made, you can use external white and dark reference files in this step. If you are reading data from ENVI `.raw` files, it may be important to use `noflip = TRUE` in your `terra::rast()` call, otherwise files will be flipped vertically.

Our test dataset is a 9 × 9 pixel, 101-band subset of HSI data (517–772 nm) from the Hyperspectral Imaging Laboratory in Gdańsk.

```{r}
# Read in captured data and references
data <- list(
  x = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/testdata.tif"
    )
  ),
  whiteref = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/WHITEREF_testdata.tif"
    )
  ),
  darkref = terra::rast(
    system.file(
      package = "HSItools",
      "testdata/capture/DARKREF_testdata.tif"
    )
  )
)
```

Reflectance is a fraction representing the proportion of light reflected by the specimen relative to the white reference. The `tint` argument controls the integration times for white reference and specimen capture, in that order. The default `c(1, 1)` assumes both use the same integration time.

## Three calibration paths

`hsi_calc_reflectance()` supports three calibration paths depending on your scanning workflow.

**Single session** is the simplest case: specimen, white reference, and dark reference all share the same integration time. No scaling is needed. This produces correct reflectance, though signal-to-noise is lower than with a dual-exposure strategy that overexposes the specimen to maximise signal.

```{r}
# Single session: equal integration times (default)
r1 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
)
```

**Matched darks** is the recommended approach for dual-exposure workflows where the specimen is scanned at a higher integration time than the white reference. Each dark subtraction uses the dark reference captured at the matching integration time. For example Lumo Scanner always captures a dark reference per session, so both are always available.

```{r}
#| eval: false
# Matched darks: recommended for dual-exposure scanning
r2 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  darkspec = data$darkspec,
  tint = c(3, 9),
  in_memory = TRUE
)
```

**Scaled dark** is a fallback for dual-exposure workflows when only the white-session dark reference is available. The dark reference is scaled by the integration time ratio before subtraction in the numerator. A warning is emitted when this path is active.

::: {.callout-warning appearance="minimal"}
Dark current scaling assumes linear scaling with integration time. In practice, numerous detectors (especially SWIR) have a large fixed-pattern noise component that does not scale with exposure time. Scaling overestimates the specimen dark current, producing severely degraded SWIR reflectance which is often negative across entire spectra. For VNIR the effect is smaller in absolute terms but we found that it introduces a systematic spectral tilt that biases indices comparing blue and NIR bands. Always prefer matched darks.
:::

```{r}
#| eval: false
# Scaled dark: fallback, not recommended
r3 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  tint = c(3, 9),
  in_memory = TRUE
)
```

## Memory and file handling

You can control whether processing is done in memory or not. With the default `in_memory = FALSE`, {[withr](https://cran.r-project.org/package=withr)} manages temporary files and removes them once the function finishes. You can write the result to a permanent file via `filename`, or keep it in memory.

```{r}
# Calculate with defaults: temporary files, result in memory
r4 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref
)

# Calculate in memory and write to a permanent file
r5 <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE,
  filename = tempfile(fileext = ".tif")
)

# Results should be nearly identical, differing only at distant decimal places
terra::all.equal(r1, r4)
```

# Smoothing: spatial

Before working in the spectral dimension, apply a spatial median filter to remove outlier pixels. The default window size is 3 × 3. As always, you can write the result to a file.

```{r}
r6 <- HSItools::hsi_smooth_median(x = r5)
```

# Smoothing: spectral

With data smoothed across the spatial dimension there should be no `NA` values in the spectral series. You can now apply a Savitzky-Golay filter to smooth the reflectance spectrum in each pixel. Several arguments are available to control the filter parameters. The function uses the {[gsignal](https://CRAN.R-project.org/package=gsignal)} implementation.

```{r}
r7 <- HSItools::hsi_smooth_savgol(x = r6)
```

## Derivatives

Derivatives of reflectance are useful for enhancing subtle spectral features and are calculated in the same function by changing the derivative order. A first derivative emphasises the rate of change in reflectance across wavelengths and can reveal absorption features that are difficult to detect in the raw spectrum. Set `m = 1` for a first derivative.

Note that edge effects at the first and last few bands (roughly equal to half the filter window width) produce unreliable derivative values. It is best to calculate derivatives on the full spectrum and subset afterwards.

```{r}
r_derivative <- HSItools::hsi_smooth_savgol(
  x = r6,
  m = 1
)
```

# Continuum removal

Smoothed data can be enhanced further by removing the continuum, the convex hull of the spectrum. This corrects for the broad slope inherent in most hyperspectral data, isolating individual absorption features and making them more comparable across pixels and samples. The function uses the {[prospectr](https://CRAN.R-project.org/package=prospectr)} implementation.

Continuum removal is computationally intensive. Consider applying it to subsets or regions of interest rather than full-resolution data.

```{r}
r8 <- HSItools::hsi_remove_continuum(x = r7)
```

# Workflow

With {HSItools} and {terra} it is straightforward to pipe operations into a single workflow. You can write to a permanent file at any step by passing a `filename` argument.

```{r}
r <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
) |>
  # Smooth spatially
  HSItools::hsi_smooth_median() |>
  # Smooth spectrally and write an intermediate checkpoint
  HSItools::hsi_smooth_savgol(filename = tempfile(fileext = ".tif")) |>
  # Remove continuum
  HSItools::hsi_remove_continuum()
```

# Tiled processing

If your computer has multiple cores you can apply pixel-wise operations in parallel using `hsi_tiled()`. Workers are managed by {[mirai](https://CRAN.R-project.org/package=mirai)}.

::: {.callout-caution appearance="minimal"}
Tiled processing produces reliable results only for operations that work on a single-pixel basis. Spatial operations such as the median filter must never be run this way, as tiles do not share border information.
:::

```{r}
library(mirai)

# Set number of parallel workers.
# Reserve at least one core for the main process.
cores <- 2
mirai::daemons(cores)
```

```{r}
r <- HSItools::hsi_calc_reflectance(
  x = data$x,
  whiteref = data$whiteref,
  darkref = data$darkref,
  in_memory = TRUE
) |>
  # Smooth spatially — no tiling
  HSItools::hsi_smooth_median() |>
  # Smooth spectrally — tiling is safe here
  HSItools::hsi_tiled(
    fun = \(tile) HSItools::hsi_smooth_savgol(tile),
    n_tiles = cores,
    filename = tempfile(fileext = ".tif"),
    overwrite = TRUE
  ) |>
  # Remove continuum — tiling is safe here
  HSItools::hsi_tiled(
    fun = \(tile) HSItools::hsi_remove_continuum(tile),
    n_tiles = cores,
    filename = tempfile(fileext = ".tif"),
    overwrite = TRUE
  )

# Shut down workers when done
mirai::daemons(0)
```

# Storage: quantizing

High spatial and spectral resolution cameras, especially VNIR, produce large reflectance files once data moves from integer DN space into floating-point numbers. One way to reduce file size is to store a *quantized* version: values are multiplied by a known scale factor and stored as integers, then divided back on read. This typically reduces file size by more than 50%.

{terra} reads quantized files transparently. The tradeoff is slightly reduced precision of the restored values, which is generally acceptable for high-SNR sensors.

::: {.callout-warning appearance="minimal"}
Do not quantize low-SNR data such as current-generation SWIR sensors. Quantization discards the fine floating-point differences that carry meaningful signal in noisy data.
:::

`hsi_write_scaled()` is a write-primary function: it encodes `x` to disk and returns the original object invisibly so it can sit in a pipeline without printing. Continue processing from the object you passed in, not from the return value.

```{r}
HSItools::hsi_write_scaled(
  x = r,
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)
```