---
title: "spatial_calibration"
vignette: >
  %\VignetteIndexEntry{spatial_calibration}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(HSItools)
```

Hyperspectral imaging data is, by definition, spatial. However, typically there is no easy and immediate anchor to real world units. Most of the time ENVI or GeoTIFF will load in GIS software in WGS84 or with NULL CRS.

# Getting the units

To express HSI data in real world units we need, at the very minimum, to know real world resolution. Internally, we use micrometers per pixel throughout the HSItools package and offer options to convert to mm and cm. We also offer three complementary ways to get and store real world resolution. We assume throughout the workflow, that pixels are square.

## Known resolution

In an instance where real world pixel size is known, for example in a lab where field of view is used to determine scanning resolution, or once one scan is done, and others follow exactly the same settings, it is possible to simply provide a number.

```{r}
um <- HSItools::hsi_calibration_direct(60, units = "um")

um
```

## Digitized scale

If there is a vertical or horizontal measure scanned along the specimen it is easy to digitize it and use as a reference. Simply read your digitized data as a `terra` SpatVector pointing to a line or two-point feature and provide length that was read out. Importantly, be sure that the CRS is NULL.

```{r}
# Construct SpatVector by hand
line_vect <- terra::vect(
  matrix(c(1000, 2000, 1000, 2100), ncol = 2, byrow = TRUE),
  type = "lines"
) |>
  HSItools::hsi_drop_crs()

# Distance of 10000 micrometers is default
um <- HSItools::hsi_calibration_from_scale(
  line_vect,
  distance = 10000,
  units = "um"
)

um
```

## Known dimensions

If you know the exact scan length that was used in the controlling software, you can use it as a robust reference. We use known scan length and number of rows in a SpatRaster to calculate resolution. This requires that kind of metadata is written down, but ensures very good resolution estimate.

```{r}
r <- terra::rast(
  system.file(
    package = "HSItools",
    "testdata/products/SAVGOL_testdata.tif"
  )
)

um <- HSItools::hsi_calibration_from_dims(
  terra::nrow(r),
  distance = 10000,
  units = "um"
)
```

# Building a coordinate raster

# Anchoring to a reference point