---
title: "Spectral indices"
vignette: >
  %\VignetteIndexEntry{spectral_indices}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(HSItools)
```

# Spectral indices catalogue

HSItools includes a built-in catalogue of spectral indices drawn from the literature. It is a practical starting point for common sediment proxies, not a comprehensive literature review.

```{r}
# Show tibble
HSItools::spectral_indices
```

You can filter by `proxy_type` to see what is available for a given calculation method:

```{r}
HSItools::spectral_indices |>
  dplyr::filter(proxy_type == "rabd")
```

# Calculating spectral indices

Given the known spectral features, we can now calculate our first spectral index of choice. To calculate an index, pass the spectral parameters directly to the corresponding function. We are going to use test reflectance previously processed with `hsi_smooth_median()` and `hsi_smooth_savgol()`. We also need a derivative raster.

```{r}
r <- terra::rast(
  system.file(
    package = "HSItools",
    "testdata/products/SAVGOL_testdata.tif"
  )
)
```

## Relative Absorption Band Depth

Relative Absorption Band Depth (RABD) is a straightforward type of index, often used for sediment scanning data. Calculate RABD centered around 660 nm to see relative chlorophyll-a abundance.

```{r}
# A "strict" variant pointing at a given wavelength
rabd660 <- HSItools::hsi_calc_rabd(
  r,
  continuum_edges = c(590, 730),
  absorption_band = 660,
  index_type = "strict",
  index_name = "rabd660",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

rabd660
```

```{r}
# A "max" variant flexibly selecting wavelength of a minimum reflectance value in a range
rabd660670 <- HSItools::hsi_calc_rabd(
  r,
  continuum_edges = c(590, 730),
  absorption_band = c(660, 670),
  index_type = "max",
  index_name = "rabd660670",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

rabd660670
```

## Band Ratio

Band ratio is a simple index where reflectance at band A is divided by reflectance at band B.

```{r}
ratio660670 <- HSItools::hsi_calc_ratio(
  r,
  bands = c(660, 670),
  index_name = "ratio660670",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

ratio660670
```

## Band Difference

Band difference subtracts reflectance at band B from reflectance at band A.

```{r}
diff660690 <- HSItools::hsi_calc_difference(
  r,
  bands = c(660, 690),
  index_name = "diff660690",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

diff660690
```

## Red-Edge Minimum Point

λREMP is an index meant for chlorophyll-a detection proposed by Ghanbari et al. (2023). It uses the first derivative of the spectrum to find exact wavelength where the absorbance near the red edge reaches minimum (zero crossing in derivative space). The wavelength (nm) is the index value. By default we are looking for minimum value (dip) in 660 nm to 680 nm.

```{r}
remp <- terra::rast(
  system.file(
    package = "HSItools",
    "testdata/products/MEDIAN_testdata.tif"
  )
) |>
  # Calculate 1st derivative
  HSItools::hsi_smooth_savgol(m = 1) |>
  # Calculate REMP
  HSItools::hsi_calc_remp(
    search_range = c(660, 680),
    filename = tempfile(fileext = ".tif"),
    index_name = "remp",
    overwrite = TRUE
  )

remp
```

# Using the catalogue

Instead of typing spectral parameters manually, you can look them up directly from `spectral_indices` and pass them to the function. This is useful when running multiple indices or when you want to keep your code tied to a single source of truth. Here we recalculate `rabd660670` using the catalogue rather than typed parameters. The result is identical.

```{r}
# Pull the rabd660670 entry
idx <- HSItools::spectral_indices |>
  dplyr::filter(proxy_name == "rabd660670")

# Pass parameters to the function
rabd660670 <- HSItools::hsi_calc_rabd(
  x = r,
  continuum_edges = idx$continuum_edges[[1]],
  absorption_band = idx$absorption_band[[1]],
  index_type = idx$index_type,
  index_name = "rabd660670",
  filename = tempfile(fileext = ".tif"),
  overwrite = TRUE
)

rabd660670
```

# Data extraction

## Extract index profile

Spatial data is an important foundation of HSI, however, at the end we are often interested in getting a single profile representing an index.

```{r}
rabd_profile <- rabd660670 |>
  HSItools::hsi_extract_profile()

rabd_profile
```

You can also easilly plot these values with a helper function. These are `ggplot2` objects that can be easily modified with `+`. Please mind, that at this moment we are using the `terra` pixel space. Spatial calibration is covered in [spatial calibration vignette](spatial_calibration.html).

::: callout-important
## The plot function requires the index raster to have a named layer. Pass `index_name` when calculating to ensure this..
:::

```{r}
rabd_profile |> HSItools::hsi_plot_profile()
```

## Extract reflectance profile

Reflectance profile in a region of choice is a good diagnostic tool. It can be easilly pulled.

```{r}
r_profile <- r |>
  HSItools::hsi_extract_spectrum()

r_profile
```

You can also easilly plot these values with a helper function, just like the `hsi_extract_profile()` output.

```{r}
r_profile |> HSItools::hsi_plot_spectrum()
```